Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-31 _tex/paper2.tex
Built site for gh-pages

A motivating example that will appear in the rest of the document
clarifies these concepts. In this example, researchers aim to determine:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this, a
researcher designs a CJ experiment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges then compare pairs of students' written texts (stimuli) to
produce a dichotomous outcome reflecting the relative quality of each
text (trait). Based on this setup, researchers can reformulate the
research question as the estimand: ``\emph{On average}, is there a
difference in the ability to produce high-quality written texts between
the two groups of students?''. Following current CJ practices,
researchers then rely on estimates from the BTL model, or its
transformations, to approximate this estimand.

@@

Although specifying a system's data structure is not mandatory when
using SCMs and DAGs, defining one in this case can improve clarity and
facilitate the description of the system. Thus, to re-express the scalar
form of the CJ system shown in Figure~\ref{fig-cj03} into an equivalent
vectorized form, we first define the vectors \(I\) and \(J\), along with
the matrices \(IA\) and \(JK\), as follows:

@@

Here, each element of \(I\) represents a unique individual \(h\) or
\(i\), where \(n_{I}\) denotes the total number of individuals.
Similarly, each element of \(J\) corresponds to a unique judge \(j\),
with \(n_{J}\) indicating the total number of judges. Moreover, each row
of \(IA\) represents a unique pairing of individuals \(h, i\) with
stimuli \(a, b\). As a result, the matrix \(IA\) contains
\((n_{I} \cdot n_{A})\) rows and \(2\) columns, where \(n_{A}\)
specifies the number of stimuli available per individual. Likewise, each
row of \(JK\) associates a judge \(j\) with a judgment index \(k\).
Consequently, the matrix \(JK\) has \((n_{J} \cdot n_{K})\) rows and
\(2\) columns, where \(n_{K}\) indicates the number of judgments each
judge makes.

@@

Now, using equations (\ref{eq-mat11}) and (\ref{eq-mat12}), we can
re-express SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an
equivalent vectorized form, as shown in Figure~\ref{fig-cj04}. In this
depiction, the outcome \(O^{cp}_{R}\), the texts' discriminal difference
\(D_{R}\), their discriminal processes \(T_{IA}\), and the judges'
biases \(B_{JK}\) are represented as vectors rather than scalar values.
These vectors capture all the observations from the conceptual
population. Specifically, \(O^{cp}_{R}\) and \(D_{R}\) are observed and
latent vectors of length \(n\), respectively. Moreover, \(T_{IA}\) and
\(B_{JK}\) are latent vectors of lengths \((n_{I} \cdot n_{A})\) and
\((n_{J} \cdot n_{K})\), respectively.

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model integrates two \emph{hierarchical
structural components} to examine how different teaching methods
influence students' writing ability. Each structural component defines
how observed or latent variables affect the primary latent variable of
interest \citep{Everitt_et_al_2010}. Their hierarchical nature enables
researchers to test hypotheses while accounting for uncertainties
associated with trait estimation (see Section~\ref{sec-theory-issue2}),
and when stimuli exhibit a hierarchical structure (see
Section~\ref{sec-theory-issue1b}).

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{Here, \emph{relevant} refers
  to variables that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. These topics are beyond the scope of this study,
  thus, readers seeking a more profound understanding can refer to
  introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
  \citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}, as well as
  introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020},
  and \citet{McElreath_2020} useful. For more advanced study, seminal
  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
  \citet{Hernan_et_al_2020}, are recommended.} student-related variables
\((X_{I})\), such as the teaching method, causally influence the latent
variable representing students' writing-quality trait \((T_{I})\). It
also shows how \(T_{I}\) and \emph{relevant} \footnote{refer to footnote
  1.} text-related variables \((X_{IA})\) (e.g., text length) causally
influence the texts' written-quality trait \((T_{IA})\), the primary
latent variable of interest. Additionally, the branch accounts for
idiosyncratic errors: \(e_{I}\) captures variations in students' traits
unexplained by \(X_{I}\), while \(e_{IA}\) captures variations in texts'
traits unexplained by \(T_{I}\) or \(X_{IA}\). Here, \(X_{I}\) is an
observed matrix with \(n_{I}\) rows and \(q_{I}\) columns (variables),
and both \(e_{I}\) and \(T_{I}\) are latent vectors of length \(n_{I}\).
Similarly, \(X_{IA}\) is an observed matrix with dimension
\(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) columns (variables), while
\(e_{IA}\) and \(T_{IA}\) are latent vectors of length
\(n_{I} \cdot n_{A}\).

@@

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 1.}
judge-related variables \((Z_{J})\), such as judgment expertise,
causally influence the latent variable representing judges' bias
\((B_{J})\). It also shows how \(B_{J}\) and \emph{relevant} \footnote{refer
  to footnote 1.} judgment-related variables \((Z_{JK})\) (e.g., the
number of judgments a judge makes) causally influence the biases
associated with each text \(B_{JK}\), the primary latent variable of
interest. Additionally, the branch accounts for idiosyncratic errors:
\(e_{J}\) captures variations in judges' bias unexplained by \(Z_{J}\),
while \(e_{JK}\) captures variations in judgments unexplained by
\(B_{J}\) or \(Z_{JK}\). Here, \(Z_{J}\) is an observed matrix with
\(n_{J}\) rows and \(q_{J}\) columns (variables), and both \(e_{J}\) and
\(B_{J}\) are latent vectors of length \(n_{J}\). Similarly, \(Z_{JK}\)
is an observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and
\(q_{JK}\) columns (variables), while \(e_{JK}\) and \(B_{JK}\) are
latent vectors of length \(n_{J} \cdot n_{K}\).

Notably, all variables and functions shown in SCM~\ref{fig-cj09_scm} and
DAG~\ref{fig-cj09_dag} define the set of endogenous variables \(V\),
structural equations \(F\), and exogenous variables \(E\) for the
conceptual-population model. Additionally, the figures demonstrate that
all exogenous variables are independent of one another, as indicated by
the relationships \(e_{IA} \:\bot\:\{ e_{I}, e_{JK}, e_{J} \}\),
\(e_{I} \:\bot\:\{ e_{JK}, e_{J} \}\) and \(e_{JK} \:\bot\:e_{J}\).

@@

Equation (\ref{eq-mat24}) shows that missing values in the \(S_{I}\) and
\(S_{J}\) vectors directly determine those in the \(S_{IA}\) and
\(S_{JK}\). In other words, researchers can only observe texts and
judgments from students and judges initially included in the sample.
Additionally, it shows that the sum of observed elements in \(S_{I}\)
equals the sampled students \((n^{s}_{I})\) and that a similar sum in
vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\). Conversely,
the sum of observed elements in \(S_{IA}\) represents the total sampled
texts across all students \((n^{s}_{I} \cdot n^{s}_{A})\), while a
similar sum in vector \(S_{JK}\) represents the total sampled judgments
across all judges \((n^{s}_{J} \cdot n^{s}_{K})\).

Finally, we define the \emph{sample design} vector \(S\) in equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
of \(S_{JK}\). The sum of its elements represents the total data sample
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).

@@

The DAG also depicts the \emph{sample design} vector \(S\) as a causal
factor influencing the sample-comparison outcome \(O^{sc}_{R}\). A
square encloses this vector, indicating that it is a conditioned
variable. In this context, \emph{conditioning} means restricting the
statistical analysis to a specific subset of the population based on the
values of a conditioned variable \citep[pp.~32]{Neal_2020}. Essentially,
\(S\) serves as the sample equivalent for the \(R\) matrix (see equation
(\ref{eq-mat12})), that is, this vector determines \emph{which repeated
comparisons a subset of judges makes for a subset of stimuli produced by
a subset of individuals} from the conceptual-population, ultimately
generating the sample-comparison outcome \(O^{sc}_{R}\) from the
conceptual-population outcome \(O^{cp}_{R}\).

@@

However, because CJ experiments rarely exhaustively pair sampled
judgments, judges, stimuli, and individuals--due to concerns about the
practical feasibility of the comparison task, like in
\citet[p.~5]{Boonen_et_al_2020}--a realistic scenario must account for
judges comparing only specific stimuli from certain individuals.

>>>>

A motivating example that will appear in the rest of the document
clarifies these concepts. In this example, researchers aim to determine:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this, a
researcher designs a CJ experiment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges then compare pairs of students' written texts (stimuli) to
produce a dichotomous outcome reflecting the relative quality of each
text (trait). Based on this setup, researchers can reformulate the
research question as the estimand: ``\emph{On average}, is there a
difference in the ability to produce high-quality written texts between
the two groups of students?''. Following current CJ practices,
researchers rely on estimates from the BTL model, or its
transformations, to approximate this estimand.

@@

Although specifying a data structure is not mandatory when using SCMs
and DAGs, defining one in this case can improve clarity and facilitate
the description of the system. Thus, to re-express the scalar form of
the CJ system shown in Figure~\ref{fig-cj03} into an equivalent
vectorized form, we first define the vectors \(I\) and \(J\), along with
the matrices \(IA\) and \(JK\), as follows:

@@

Here, each element of \(I\) represents a unique individual \(h\) or
\(i\), where \(n_{I}\) denotes the total number of individuals.
Similarly, each element of \(J\) corresponds to a unique judge \(j\),
with \(n_{J}\) indicating the total number of judges. Moreover, each row
of \(IA\) represents a unique pairing of individuals \(h, i\) with
stimuli \(a, b\). As a result, the matrix \(IA\) contains
\(n_{I} \cdot n_{A}\) rows and \(2\) columns, where \(n_{A}\) specifies
the number of stimuli available per individual. Likewise, each row of
\(JK\) associates a judge \(j\) with a judgment index \(k\).
Consequently, the matrix \(JK\) has \(n_{J} \cdot n_{K}\) rows and \(2\)
columns, where \(n_{K}\) indicates the number of judgments each judge
makes.

@@

Now, using equations (\ref{eq-mat11}) and (\ref{eq-mat12}), we can
re-express SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an
equivalent vectorized form, as shown in Figure~\ref{fig-cj04}. In this
depiction, the outcome \(O^{cp}_{R}\), the texts' discriminal difference
\(D_{R}\), their discriminal processes \(T_{IA}\), and the judges'
biases \(B_{JK}\) are represented as vectors rather than scalar values.
These vectors capture all the observations from the conceptual
population. Specifically, \(O^{cp}_{R}\) and \(D_{R}\) are observed and
latent vectors of length \(n\), respectively. Moreover, \(T_{IA}\) and
\(B_{JK}\) are latent vectors of lengths \(n_{I} \cdot n_{A}\) and
\(n_{J} \cdot n_{K}\), respectively.

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model integrates two \emph{hierarchical
structural components} to examine how different teaching methods
influence students' writing ability. Each structural component defines
how observed or latent variables affect the primary latent variable of
interest \citep{Everitt_et_al_2010}. Their hierarchical nature enables
researchers to test hypotheses considering the hierarchical structure of
stimuli (see Section~\ref{sec-theory-issue1b}) and the uncertainties in
trait estimation (see Section~\ref{sec-theory-issue2}).

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{Here, \emph{relevant} refers
  to variables that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. These topics are beyond the scope of this study,
  thus, readers seeking a more profound understanding can refer to
  introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
  \citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}, as well as
  introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020},
  and \citet{McElreath_2020} useful. For more advanced study, seminal
  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
  \citet{Hernan_et_al_2020}, are recommended.} student-related variables
\((X_{I})\), such as the teaching method, causally influence the latent
variable representing students' writing-quality trait \((T_{I})\). It
also shows how \(T_{I}\) and \emph{relevant} \footnote{refer to footnote
  1.} text-related variables \((X_{IA})\) (e.g., text length) causally
influence the texts' written-quality trait \((T_{IA})\), the first
primary latent variable of interest. Additionally, the branch accounts
for idiosyncratic errors: \(e_{I}\) captures variations in students'
traits unexplained by \(X_{I}\), while \(e_{IA}\) captures variations in
texts' traits unexplained by \(T_{I}\) or \(X_{IA}\). Here, \(X_{I}\) is
an observed matrix with \(n_{I}\) rows and \(q_{I}\) columns
(variables), and both \(e_{I}\) and \(T_{I}\) are latent vectors of
length \(n_{I}\). Similarly, \(X_{IA}\) is an observed matrix with
dimension \(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) columns (variables),
while \(e_{IA}\) and \(T_{IA}\) are latent vectors of length
\(n_{I} \cdot n_{A}\).

@@

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 1.}
judge-related variables \((Z_{J})\), such as judgment expertise,
causally influence the latent variable representing judges' bias
\((B_{J})\). It also shows how \(B_{J}\) and \emph{relevant} \footnote{refer
  to footnote 1.} judgment-related variables \((Z_{JK})\) (e.g., the
number of judgments a judge makes) causally influence the biases
associated with each text \(B_{JK}\), the second primary latent variable
of interest. Additionally, the branch accounts for idiosyncratic errors:
\(e_{J}\) captures variations in judges' bias unexplained by \(Z_{J}\),
while \(e_{JK}\) captures variations in judgments unexplained by
\(B_{J}\) or \(Z_{JK}\). Here, \(Z_{J}\) is an observed matrix with
\(n_{J}\) rows and \(q_{J}\) columns (variables), and both \(e_{J}\) and
\(B_{J}\) are latent vectors of length \(n_{J}\). Similarly, \(Z_{JK}\)
is an observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and
\(q_{JK}\) columns (variables), while \(e_{JK}\) and \(B_{JK}\) are
latent vectors of length \(n_{J} \cdot n_{K}\).

Notably, all variables and functions shown in SCM~\ref{fig-cj09_scm} and
DAG~\ref{fig-cj09_dag} are part of the set of endogenous variables
\(V\), structural equations \(F\), and exogenous variables \(E\) for the
conceptual-population model. Additionally, the figures demonstrate that
all exogenous variables are independent of one another, as indicated by
the relationships \(e_{IA} \:\bot\:\{ e_{I}, e_{JK}, e_{J} \}\),
\(e_{I} \:\bot\:\{ e_{JK}, e_{J} \}\) and \(e_{JK} \:\bot\:e_{J}\).

@@

Equation (\ref{eq-mat24}) shows that missing values in the \(S_{I}\) and
\(S_{J}\) vectors directly determine those in the \(S_{IA}\) and
\(S_{JK}\). In other words, researchers can only observe texts and
judgments from students and judges initially included in the sample.
Additionally, it shows that the sum of observed elements in \(S_{I}\)
equals the sampled students \((n^{s}_{I})\) and that a similar sum in
vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\). Conversely,
the sum of observed elements in \(S_{IA}\) represents the total sampled
texts across all sampled students \((n^{s}_{I} \cdot n^{s}_{A})\), while
a similar sum in vector \(S_{JK}\) represents the total sampled
judgments across all sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\).

Finally, we define the \emph{sample design} vector \(S\) in equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
of \(S_{JK}\). The sum of its elements represents the total data sample
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
binomial coefficient, which quantifies the total number of unique
comparisons possible between every pair of sampled stimuli generated by
each pair of sampled individuals from the population.

@@

The DAG also depicts the \emph{sample design} vector \(S\) as a causal
factor influencing the sample-comparison outcome \(O^{sc}_{R}\). A
square encloses this vector, indicating that it is a conditioned
variable. In this context, \emph{conditioning} means restricting the
statistical analysis to a specific subset of the population based on the
values of a conditioned variable \citep[pp.~32]{Neal_2020}, in this case
\(S\). Essentially, \(S\) serves as the sample equivalent for the \(R\)
matrix (see equation (\ref{eq-mat12})), that is, this vector determines
\emph{which repeated comparisons the subset of judges makes for the
subset of stimuli produced by the subset of individuals} sampled from
the conceptual-population. This process ultimately generates the
sample-comparison outcome \(O^{sc}_{R}\) from the conceptual-population
outcome \(O^{cp}_{R}\).

@@

However, because CJ experiments rarely use the exhaustive pairing of the
sampled judgments, judges, stimuli, and individuals --due to concerns
about the practical feasibility of the comparison task, like in
\citet[p.~5]{Boonen_et_al_2020}-- a realistic scenario must also account
for judges comparing only specific stimuli from certain individuals.
